---
title: Is parking the bus worth it?
date: '2017-09-28'
slug: balance
categories:
  - football
  - statistics
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(readxl)
premier_league_historical_stats <- 
  read_excel("premier_league_historical_stats.xlsx") %>% 
  filter(league == "England") %>% 
  select(team, season, points, goals_for, goals_against) %>% 
  arrange(season, desc(points))
```

There has been an ongoing discussion on Twitter on whether it is better for
underdogs to play defensively or offensively. 

```{r echo=FALSE}
blogdown::shortcode('tweet', '912944800730624000')
```

Now, to me, it is a mathematical certainty that lower-scoring games are more 
random, and thus better for the underdog. Conversely, higher-scoring games are 
better for the favorites, who are more assured of winning as more goals lead
to less random outcomes.

It is a pretty simple point, and it squares with the general perceptions people
have about the football. Big teams play more defensive setups when playing
against other big teams. Games against defensive-minded minnows like Burnley are 
feared more than those against attack-minded sides like Bournemouth. Most people 
who watch football intuitively feel like a team has got to defend when playing
against a better team.

Most, but not all.

As every other claim made about football, this one also draws its fair share of
resistance and pushback. People don't always believe it, and they try to back
their case up with data. Frequently, a lack of mathematical and statistical
knowledge leads to mishandling of data, especially in football where we are
dealing with small amounts of count data.

So I decided to show, using real data, how it is better for relegation 
contenders to be defensive and for title contenders to be attacking. I hope this 
empirical exercise will convince people that it really is an underlying 
mathematical feature we are dealing with here, even if they can't fully 
understand the mathematics that go into it.

Our dataset contains 15 Premier League seasons, with points and goals.

```{r dataset, echo=FALSE}
premier_league_historical_stats
```

The goal is to model total points as a function of the goals scored and 
conceded. First we have to re-parameterize goals scored/conceded to an 
unconstrained scale, as we are going to use them in a linear model later. Count
data such as goals is lower-bounded at zero, but there should be no limit to how
bad a team's attack is. We then calculate a measure of quality and a measure of 
attacking/defensive balance.

```{r reparam}
premier_league_historical_stats <- 
  premier_league_historical_stats %>% 
  mutate(
    
    # First, see how many goals the team scores relative to the average
    norm_attack = goals_for %>% divide_by(mean(goals_for)) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    # First, see how many goals the team concedes relative to the average
    norm_defense = goals_against %>% divide_by(mean(goals_against)) %>% 
      # Invert it, so a higher defense is better
      raise_to_power(-1) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    
    # Now that we have normalized attack and defense ratings, we can compute
    # measures of quality and attacking balance
    
    quality = norm_attack + norm_defense,
    balance = norm_attack - norm_defense
  )
```

We now have two metrics, `quality` and `balance`, that capture how good and
how attacking a team is, respectively. Incidentally, `quality` is exactly
the log of the goal ratio, a metric that is familiar to football fans. 
`balance` is it's lesser-kown sibling, that we can christen as *goal product*
as it should map directly onto the product of the goals scored and the goals
conceded.

These are the most attacking teams in the past 15 seasons:

```{r attacking, echo=FALSE}
premier_league_historical_stats %>% 
  arrange(desc(balance)) %>% 
  select(team:goals_against, balance)
```

And these are the most defensive:

```{r defensive, echo=FALSE}
premier_league_historical_stats %>% 
  arrange(balance) %>% 
  select(team:goals_against, balance)
```

Now we can fit a simple linear model of points, where the predictors are:

+ Quality: the greater the quality, the higher the total points
+ Interaction between quality and balance: according to our "hypothesis", for 
high-quality teams, an attacking balance should lead to more points. For 
low-quality teams, the reverse should be true.

```{r model}
fit <- 
  premier_league_historical_stats %>% 
  lm(points ~ quality + quality:balance, data = .)

summary(fit)
```

I've shown the p-values above, in case you care about them (which you shouldn't,
but horses for courses, etc).

So we would expect a team of average quality and average balance to get about
52 points over a season. The team quality obviously has a huge impact on the
team's point return.

Most importantly, we see that the interaction term `quality:balance` is 
positive, meaning when both quality and balance are positive (i.e. attacking 
title contenders) a team will get more points and that when both are negative
(i.e. defensive relegation battlers) it will also get more points.

This hopefully should prove our hypothesis. However, just to make sure, let's
look at it from a different angle. Let's fit one model to attacking teams
and another to defensive teams and see how they differ. We are expecting 
attacking teams' points totals to display a stronger dependence on the team
quality. In other words, defensive teams tend towards an average points return,
whereas attacking teams tend towards an extreme (good or bad) points return.

Results in plot form:

```{r model_plot}
premier_league_historical_stats %>% 
  ggplot(aes(quality, points, fill = balance)) +
  geom_smooth(data = . %>% filter(balance > 0),
    method = "lm", color = scales::muted("red"), fill = scales::muted("red")
  ) +
  geom_smooth(data = . %>% filter(balance < 0),
    method = "lm", color = scales::muted("blue"), fill = scales::muted("blue")
  ) +
  geom_point(pch = 21) +
  scale_fill_gradient2(
    low = scales::muted("blue"), 
    high = scales::muted("red"), 
    midpoint = 0
  ) +
  theme_classic()
```

As we can see, the red line, fit on the attacking teams, has a greater slope 
than the blue line. The red dots are higher than the blue dots at high quality, 
but lower at low quality.

**For good teams, it is better to be offensive, and for bad teams, it is better
to be defensive**

### Bundesliga

Because I'm absolutely sure it will come up, I've gone through the same process
for the Bundesliga:

```{r bl_setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(readxl)
bundesliga_historical_stats <- 
  read_excel("premier_league_historical_stats.xlsx") %>% 
  filter(league == "Germany") %>% 
  select(team, season, points, goals_for, goals_against) %>% 
  arrange(season, desc(points)) %>% 
  mutate(
    
    # First, see how many goals the team scores relative to the average
    norm_attack = goals_for %>% divide_by(mean(goals_for)) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    # First, see how many goals the team concedes relative to the average
    norm_defense = goals_against %>% divide_by(mean(goals_against)) %>% 
      # Invert it, so a higher defense is better
      raise_to_power(-1) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    
    # Now that we have normalized attack and defense ratings, we can compute
    # measures of quality and attacking balance
    
    quality = norm_attack + norm_defense,
    balance = norm_attack - norm_defense
  )
```

```{r bundesliga_model, echo=FALSE}
fit <- 
  bundesliga_historical_stats %>% 
  lm(points ~ quality + quality:balance, data = .)

summary(fit)
```

```{r bundesliga_plot, echo=FALSE}
bundesliga_historical_stats %>% 
  ggplot(aes(quality, points, fill = balance)) +
  geom_smooth(data = . %>% filter(balance > 0),
    method = "lm", color = scales::muted("red"), fill = scales::muted("red")
  ) +
  geom_smooth(data = . %>% filter(balance < 0),
    method = "lm", color = scales::muted("blue"), fill = scales::muted("blue")
  ) +
  geom_point(pch = 21) +
  scale_fill_gradient2(
    low = scales::muted("blue"), 
    high = scales::muted("red"), 
    midpoint = 0
  ) +
  theme_classic()
```

### La Liga

I've also done La Liga for good measure:

```{r ll_setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(readxl)
la_liga_historical_stats <- 
  read_excel("premier_league_historical_stats.xlsx") %>% 
  filter(league == "Spain") %>% 
  select(team, season, points, goals_for, goals_against) %>% 
  arrange(season, desc(points)) %>% 
  mutate(
    
    # First, see how many goals the team scores relative to the average
    norm_attack = goals_for %>% divide_by(mean(goals_for)) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    # First, see how many goals the team concedes relative to the average
    norm_defense = goals_against %>% divide_by(mean(goals_against)) %>% 
      # Invert it, so a higher defense is better
      raise_to_power(-1) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    
    # Now that we have normalized attack and defense ratings, we can compute
    # measures of quality and attacking balance
    
    quality = norm_attack + norm_defense,
    balance = norm_attack - norm_defense
  )
```

```{r la_liga_model, echo=FALSE}
fit <- 
  la_liga_historical_stats %>% 
  lm(points ~ quality + quality:balance, data = .)

summary(fit)
```

```{r la_liga_plot, echo=FALSE}
la_liga_historical_stats %>% 
  ggplot(aes(quality, points, fill = balance)) +
  geom_smooth(data = . %>% filter(balance > 0),
    method = "lm", color = scales::muted("red"), fill = scales::muted("red")
  ) +
  geom_smooth(data = . %>% filter(balance < 0),
    method = "lm", color = scales::muted("blue"), fill = scales::muted("blue")
  ) +
  geom_point(pch = 21) +
  scale_fill_gradient2(
    low = scales::muted("blue"), 
    high = scales::muted("red"), 
    midpoint = 0
  ) +
  theme_classic()
```

Here the two fits don't completely separate at the lower end of the scale.
Simpson's paradox and all that. Remember that what we trust the linear fit 
above better. Fitting one model for attacking teams and another for defensive
teams is a very clumsy way to answer the question, but I thought it would at
least be easy for people to understand.

The model coefficients are nearly identical to the Premier League ones. I 
truly, honestly hope there are no "It's different for La Liga" takes, funny
though it would be.

### All together

Putting all three leagues together (normalizing Bundesliga points):

```{r all_setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(readxl)
historical_stats <- 
  read_excel("premier_league_historical_stats.xlsx") %>% 
  mutate(points = if_else(league == "Germany", points * 38/34, points)) %>% 
  select(team, season, points, goals_for, goals_against) %>% 
  arrange(season, desc(points)) %>% 
  mutate(
    # First, see how many goals the team scores relative to the average
    norm_attack = goals_for %>% divide_by(mean(goals_for)) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    # First, see how many goals the team concedes relative to the average
    norm_defense = goals_against %>% divide_by(mean(goals_against)) %>% 
      # Invert it, so a higher defense is better
      raise_to_power(-1) %>% 
      # Then, transform it into an unconstrained scale
      log(),
    
    # Now that we have normalized attack and defense ratings, we can compute
    # measures of quality and attacking balance
    
    quality = norm_attack + norm_defense,
    balance = norm_attack - norm_defense
  )
```

```{r all_model, echo=FALSE}
fit <- 
  la_liga_historical_stats %>% 
  lm(points ~ quality + quality:balance, data = .)

summary(fit)
```

```{r all_plot, echo=FALSE}
historical_stats %>% 
  ggplot(aes(quality, points, fill = balance)) +
  geom_point(pch = 21, alpha = 0.5) +
  geom_smooth(data = . %>% filter(balance > 0),
    method = "lm", color = scales::muted("red"), fill = scales::muted("red")
  ) +
  geom_smooth(data = . %>% filter(balance < 0),
    method = "lm", color = scales::muted("blue"), fill = scales::muted("blue")
  ) +
  scale_fill_gradient2(
    low = scales::muted("blue"), 
    high = scales::muted("red"), 
    midpoint = 0
  ) +
  theme_classic()
```

### Conclusion

This was interesting, but it should not be necessary. Whether to be offensive or
defensive is one of the most basic mathematical questions one can ask about
football; it is "Mathematics in Football 101". It is dispiriting that
this should be a matter of confusion. If football analysts can't even answer 
this simple question, how could they possibly tackle more complex questions?

Football has little data. Goals are infrequent, and games in different 
competitions are not immediately comparable. When the data is sparse, it is
imperative to use proper mathematical and statistical methodology. Rules of 
thumb won't do it, and disregard for the nature of the measurements (i.e.
multiplicative count data vs. linearized dimensionless quantities) is 
disastrous. Humans have spent thousands of years developing mathematics - the 
least we can do is learn it and use it.
